---
title: "Getting Started with hhdynamics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with hhdynamics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**hhdynamics** fits a Bayesian household transmission model using MCMC to estimate:

- The daily probability of infection from the community
- The probability of person-to-person transmission within households
- Effects of covariates on infectivity and susceptibility

The model accounts for tertiary transmission (household contacts infecting other household contacts) and uses a serial interval distribution to link infection times.

## Data format

Input data must be in long format (one row per individual) with these required columns:

| Column   | Description |
|----------|-------------|
| `hhID`   | Household identifier |
| `member` | Member index within household (0 = index case, 1, 2, ... = contacts) |
| `size`   | Number of individuals in the household |
| `end`    | End date of follow-up for that individual |
| `inf`    | Infection status (1 = infected, 0 = not infected). Index cases must be infected. |
| `onset`  | Onset time of symptoms (use -1 or NA for uninfected) |

Additional columns can be included as covariates for infectivity or susceptibility.

```{r data}
library(hhdynamics)
data("inputdata")
str(inputdata)
head(inputdata)
```

## Serial interval

The model requires a discrete serial interval distribution as a probability mass function of length 14:

```{r si}
data("SI")
print(SI)
barplot(SI, names.arg = 1:14, xlab = "Days", ylab = "Probability",
        main = "Serial interval distribution")
```

## Fitting the model

Use `household_dynamics()` to fit the model. Specify covariates using R formulas:

```{r fit, eval = FALSE}
# With covariates
fit <- household_dynamics(inputdata, inf_factor = ~sex, sus_factor = ~age, SI = SI,
                          n_iteration = 50000, burnin = 10000, thinning = 10)

# Without covariates
fit <- household_dynamics(inputdata, SI = SI,
                          n_iteration = 50000, burnin = 10000, thinning = 10)
```

The function returns an `hhdynamics_fit` object containing the full MCMC output.

## Inspecting results

### Quick overview

```{r print, eval = FALSE}
print(fit)
# Household transmission model fit
#   Data: 386 households, 1533 individuals
#   MCMC: 4000 post-burnin samples (50000 iterations, burnin: 10000, thin: 10)
#   Runtime: 85 seconds
#   Parameters: community, household, sex1.0, age1.0, age2.0
#
# Use summary() for estimates, coef() for posterior means.
```

### Parameter estimates

`summary()` returns a table with posterior means, 95% credible intervals, and exponentiated estimates for covariate effects:

```{r summary, eval = FALSE}
summary(fit)
#                                                    Variable Point estimate Lower bound Upper bound ...
#               Daily probability of infection from community          0.004       0.002       0.007
#  Probability of person-to-person transmission in households          0.057       0.034       0.084
#                                                      sex1.0        -0.081      -0.733       0.488
#                                                      age1.0        -0.065      -0.537       0.412
#                                                      age2.0        -0.312      -0.831       0.170
```

Community and household parameters are reported on the probability scale (via `1 - exp(-x)` transform). Covariate effects are on the log scale, with exponentiated values also shown.

### Posterior means

```{r coef, eval = FALSE}
coef(fit)
#        re_sd    community    household   size_param       sex1.0       age1.0       age2.0
#  1.000000000  0.004239978  0.058555948  0.000000000 -0.080689680 -0.065001794 -0.312414284
```

## Accessing MCMC output

The fit object stores all MCMC output for custom diagnostics and post-processing:

```{r access, eval = FALSE}
# Posterior samples matrix (post-burnin, thinned)
dim(fit$samples)       # e.g. 4000 x 7

# Log-likelihood trace (full chain, for convergence checks)
dim(fit$log_likelihood) # e.g. 50000 x 3

# Per-parameter acceptance rates
fit$acceptance

# Trace plot for community parameter
plot(fit$samples[, "community"], type = "l",
     ylab = "Community rate", xlab = "Iteration")

# Posterior density
hist(fit$samples[, "household"], breaks = 50, probability = TRUE,
     main = "Household transmission probability (raw scale)",
     xlab = "Rate parameter")
```

## Input validation

`household_dynamics()` validates inputs before running the MCMC, catching common errors early:

```{r validation, eval = FALSE}
# Missing column
household_dynamics(inputdata[, -1], SI = SI)
# Error: 'input' is missing required columns: hhID

# Wrong formula variable
household_dynamics(inputdata, inf_factor = ~nonexistent, SI = SI)
# Error: Variables in inf_factor not found in input: nonexistent

# Old string syntax (no longer supported)
household_dynamics(inputdata, inf_factor = "~sex", SI = SI)
# Error: 'inf_factor' must be a formula (e.g. ~sex) or NULL.
```

## Simulation

`simulate_data()` generates synthetic datasets for validation or power analysis:

```{r simulate, eval = FALSE}
# Use fitted parameter values
para <- coef(fit)

# Simulate 10 replicates of the original data structure
sim <- simulate_data(inputdata, rep_num = 10, inf_factor = ~sex, sus_factor = ~age,
                     SI = SI, para = para, with_rm = 0)
```
